% =============================================================================
\section{Field operators and restricted basis}
% =============================================================================

Multimode fields are described by operators $\Psiop_j^{\dagger}(\xvec)$ and $\Psiop_j(\xvec)$, where $\Psiop_j^{\dagger}$ creates a bosonic atom of spin $j$, $j = 1 \ldots C$ at location $\xvec$, and $\Psiop_j$ destroys one.
Same as with functions of coordinates, we will use the same scheme, abbreviating $\Psiop_j \equiv \Psiop_j(\xvec)$ and $\Psiop_j^\prime \equiv \Psiop_j(\xvec^\prime)$.
The commutators are
\begin{eqn}
\label{eqn:formalism:func-operators:commutators}
	[ \Psiop_j, \Psiop_k^{\prime\dagger} ]
	= \delta_{jk} \delta(\xvec^\prime-\xvec).
\end{eqn}
Field operators have type $\Psiop_j \in (\mathbb{R} \rightarrow \mathbb{H}_j)$. \todo{As far as I understand $\mathbb{H}_1 \oplus \ldots \oplus \mathbb{H}_C \equiv \mathbb{H}$, where $\rho \in \mathbb{H}$.}
Field operators can be decomposed using a single-particle basis \todo{explanation needed?}:
\begin{eqn}
	\Psiop_j = \sum_{\nvec \in \fullbasis_j} \phi_{j,\nvec} \hat{a}_{j,\nvec}.
\end{eqn}
Note that each component can have its own basis.
Single mode operators $\hat{a}_{j,\nvec}$ obey commutation relations~\eqnref{formalism:mm-aux:commutators},
the pair $j,\nvec$ serving as a mode identifier.

We want to restrict ourselves only to a subset of modes from full basis for each component: $\restbasis_j \subset \fullbasis_j$ \todo{Or $\subseteq$ if we allow infinite $\restbasis$? This will lead to some mathematical difficulties}.
New restricted field operators are
\begin{eqn}
	\Psiop_j = \sum_{\nvec \in \restbasis_j} \phi_{j,\nvec} \hat{a}_{j,\nvec}.
\end{eqn}
They map coordinates to a restricted Hilbert subspaces: $\Psiop_j \in \mathbb{H}_{\restbasis_j}$.
Because of the restricted nature of these operators, commutation relations~\eqnref{formalism:func-operators:commutators} no longer apply.
The following ones should be used instead:
\begin{eqn}
\label{eqn:formalism:func-operators:restricted-commutators}
	\left[ \Psiop_j, \Psiop_k^\prime \right]
	& = \left[ \Psiop_j^\dagger, \Psiop_k^{\prime\dagger} \right] = 0, \\
	\left[ \Psiop_j, \Psiop_k^{\prime\dagger} \right]
	& = \delta_{jk} \delta_{\restbasis_j}(\xvec^\prime, \xvec).
\end{eqn}

Let us now find the expression for high-order commutators of restricted field operators, analogous to \lmmref{formalism:mm-aux:high-order-commutators} for single-mode operators.
It can be done using the similar recursive procedure.

\begin{lemma}
	For $\Psiop \in \mathbb{H}_{\restbasis}$
	\begin{eqn*}
		\left[ \Psiop, ( \Psiop^{\prime\dagger} )^l \right]
		& = l \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}, \\
		\left[ \Psiop^\dagger, ( \Psiop^\prime )^l \right]
		& = - l \delta_{\restbasis}^* (\xvec^\prime, \xvec) ( \Psiop^\prime )^{l-1}.
	\end{eqn*}
\end{lemma}
\begin{proof}
Given that we know the expression for $\left[ \Psiop, ( \Psiop^{\prime\dagger} )^{l-1} \right]$,
the commutator of higher order can be expanded as
\begin{eqn}
	\left[ \Psiop, ( \Psiop^{\prime\dagger} )^l \right]
	& = \Psiop ( \Psiop^{\prime\dagger} )^l - ( \Psiop^{\prime\dagger} )^l \Psiop \\
	& = (
		\delta_{\restbasis} (\xvec^\prime, \xvec) + \Psiop^{\prime\dagger} \Psiop
	) ( \Psiop^{\prime\dagger} )^{l-1}
	- ( \Psiop^{\prime\dagger} )^l \Psiop \\
	& = \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}
	+ \Psiop^{\prime\dagger} (
		\Psiop ( \Psiop^{\prime\dagger} )^{l-1}
		- ( \Psiop^{\prime\dagger} )^{l-1} \Psiop
	) \\
	& = \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}
	+ \Psiop^{\prime\dagger} [
		\Psiop, ( \Psiop^{\prime\dagger} )^{l-1}
	].
\end{eqn}
Now we can get the commutator of any order starting from the known relation~\eqnref{formalism:func-operators:restricted-commutators}.
\end{proof}

A further generalisation of these relations is
\begin{lemma}
\label{lmm:formalism:func-operators:functional-commutators}
	For $\Psiop \in \mathbb{H}_{\restbasis}$
	\begin{eqn*}
		\left[ \Psiop, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
		& = \delta_{\restbasis} (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^{\prime\dagger}}, \\
		\left[ \Psiop^\dagger, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
		& = -\delta_{\restbasis}^* (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^\prime},
	\end{eqn*}
	where $f(x, y)$ is a function that can be expanded in the power series of $x$ and $y$.
\end{lemma}
\begin{proof}
Let us prove the first relation; the procedure for the second one is the same.
Without loss of generality, we can assume that $f(\Psiop^\prime, \Psiop^{\prime\dagger})$ can be expanded in power series of normally ordered operators (otherwise we can just use commutation relations).
Thus
\begin{eqn}
	\left[ \Psiop, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
	& = \sum_{r,s} f_{rs} [ \Psiop, (\Psiop^{\prime\dagger})^r (\Psiop^\prime)^s ] \\
	& = \sum_{r,s} f_{rs} [ \Psiop, (\Psiop^{\prime\dagger})^r ] (\Psiop^\prime)^s \\
	& = \sum_{r,s} f_{rs} r \delta_P(\xvec^\prime, \xvec)
		(\Psiop^{\prime\dagger})^{r-1} (\Psiop^\prime)^s \\
	& = \delta_P (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^{\prime\dagger}}.
	\qedhere
\end{eqn}
\end{proof}
