% =============================================================================
\section{Field operator calculus}
% =============================================================================

Field operators can simplify both the analysis of a master equation and its transformation to an \abbrev{fpe} using phase-space methods, which is the main topic of this thesis.
In this section we are going to outline the field operator calculus.
It is quite similar to the calculus of functional operators, described in \appref{func-calculus}.
The similarity and interconnection of field operators and functional operators will become even more evident during the description of the functional Wigner transformation in \charef{wigner}.

A multimode field of a $C$-component \abbrev{bec} in $D$ effective dimensions is described by field operators $\Psiopf_j^{\dagger}(\xvec)$ and $\Psiopf_j(\xvec)$, where $\Psiopf_j^{\dagger}$ creates a bosonic atom of spin $j$, $j = 1 \ldots C$ at a location defined by a $D$-component coordinate vector $\xvec$, and $\Psiopf_j$ destroys one.
We will use the same scheme as with functions of coordinates in \appref{func-calculus}, abbreviating $\Psiopf_j \equiv \Psiopf_j(\xvec)$ and $\Psiopf_j^\prime \equiv \Psiopf_j(\xvec^\prime)$.

The field operators obey standard bosonic commutation relations
\begin{eqn}
\label{eqn:wigner:op-calculus:commutators}
    [ \Psiopf_j, \Psiopf_k^{\prime} ]
    & = [ \Psiopf_j^\dagger, \Psiopf_k^{\prime\dagger} ]
    = 0, \\
    [ \Psiopf_j, \Psiopf_k^{\prime\dagger} ]
    & = \delta_{jk} \delta(\xvec^\prime-\xvec).
\end{eqn}
Field operators have type $\Psiopf_j \in (\mathbb{R}^D \rightarrow \mathbb{H}_j) \equiv \mathbb{FH}_j$, where the component Hilbert spaces $\mathbb{H}_j$ constitute the system Hilbert space $\mathbb{H} = \bigotimes_{j=1}^C \mathbb{H}_j$.
Field operators can be decomposed using a single-particle orthonormal basis (see~\eqnref{func-calculus:basis} for details):
\begin{eqn}
\label{eqn:wigner:op-calculus:field}
    \Psiopf_j(\xvec) = \sum_{\nvec \in \fullbasis_j} \phi_{j,\nvec}(\xvec) \hat{a}_{j,\nvec}.
\end{eqn}
Note that each component can have its own basis.
Single mode operators $\hat{a}_{j,\nvec}$ obey commutation relations~\eqnref{mm-wigner:mm:commutators}, with the pair $j,\nvec$ serving as a mode identifier.
From now on, we drop the coordinate argument of mode functions $\phi(\xvec)$.

In practice, one cannot use an infinitely sized basis in numerical calculations; some subset of modes is always chosen.
In order to take this into account, we will restrict ourselves to a subset of modes from the full basis for each component: $\restbasis_j \subset \fullbasis_j$.
The new restricted field operators are
\begin{eqn}
\label{eqn:wigner:op-calculus:restricted-field}
    \Psiop_j = \sum_{\nvec \in \restbasis_j} \phi_{j,\nvec} \hat{a}_{j,\nvec}.
\end{eqn}
These map coordinates to restricted Hilbert subspaces: $\Psiop_j \in (\mathbb{R} \rightarrow \mathbb{H}_{\restbasis_j}) = \mathbb{FH}_{\restbasis_j}$.
Because of the restricted nature of these operators, the commutation relations~\eqnref{wigner:op-calculus:commutators} no longer apply.
The following ones must be used instead:
\begin{eqn}
\label{eqn:wigner:op-calculus:restricted-commutators}
    \left[ \Psiop_j, \Psiop_k^\prime \right]
    & = \left[ \Psiop_j^\dagger, \Psiop_k^{\prime\dagger} \right] = 0, \\
    \left[ \Psiop_j, \Psiop_k^{\prime\dagger} \right]
    & = \delta_{jk} \delta_{\restbasis_j}(\xvec^\prime, \xvec),
\end{eqn}
where $\delta_{\restbasis_j}$ is the restricted delta function introduced in \appref{func-calculus}, \defref{func-calculus:restricted-delta}.

Let us now find the expression for high-order commutators of restricted field operators, analogous to the one for single-mode operators, which can be found in the book by Louisell~\cite{Louisell1990}.
For simplicity we will consider the case of a single component and, consequently, omit the component subscript of field operators in the following two lemmas.

\begin{lemma}
\label{lmm:wigner:op-calculus:moment-commutators}
    For $\Psiop \in \mathbb{FH}_{\restbasis}$,
    \begin{eqn*}
        \left[ \Psiop, ( \Psiop^{\prime\dagger} )^l \right]
        & = l \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}, \\
        \left[ \Psiop^\dagger, ( \Psiop^\prime )^l \right]
        & = - l \delta_{\restbasis}^* (\xvec^\prime, \xvec) ( \Psiop^\prime )^{l-1}.
    \end{eqn*}
\end{lemma}
\begin{proof}
Proved by induction.
Given that we know the expression for $\left[ \Psiop, ( \Psiop^{\prime\dagger} )^{l-1} \right]$,
the commutator of higher order can be expanded as
\begin{eqn}
    \left[ \Psiop, ( \Psiop^{\prime\dagger} )^l \right]
    & = \Psiop ( \Psiop^{\prime\dagger} )^l - ( \Psiop^{\prime\dagger} )^l \Psiop \\
    & = (
        \delta_{\restbasis} (\xvec^\prime, \xvec) + \Psiop^{\prime\dagger} \Psiop
    ) ( \Psiop^{\prime\dagger} )^{l-1}
    - ( \Psiop^{\prime\dagger} )^l \Psiop \\
    & = \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}
    + \Psiop^{\prime\dagger} (
        \Psiop ( \Psiop^{\prime\dagger} )^{l-1}
        - ( \Psiop^{\prime\dagger} )^{l-1} \Psiop
    ) \\
    & = \delta_{\restbasis} (\xvec^\prime, \xvec) ( \Psiop^{\prime\dagger} )^{l-1}
    + \Psiop^{\prime\dagger} [
        \Psiop, ( \Psiop^{\prime\dagger} )^{l-1}
    ].
\end{eqn}
A similar result holds for the second relation, on conjugating the first one.
Now we can get the commutator of any order starting from the known relation~\eqnref{wigner:op-calculus:restricted-commutators}.
\end{proof}

A further generalisation of these relations is
\begin{lemma}
\label{lmm:wigner:op-calculus:functional-commutators}
    For $\Psiop \in \mathbb{FH}_{\restbasis}$,
    \begin{eqn*}
        \left[ \Psiop, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
        & = \delta_{\restbasis} (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^{\prime\dagger}}, \\
        \left[ \Psiop^\dagger, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
        & = -\delta_{\restbasis}^* (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^\prime},
    \end{eqn*}
    where $f(x, y)$ is a function that can be expanded in the power series of $x$ and $y$.
\end{lemma}
\begin{proof}
Let us prove the first relation.
The procedure for the second one is the same.
Without loss of generality, we can assume that $f(\Psiop^\prime, \Psiop^{\prime\dagger})$ can be expanded in the power series of normally ordered operators (otherwise we can just use the commutation relations).
Thus,
\begin{eqn}
    \left[ \Psiop, f( \Psiop^\prime, \Psiop^{\prime\dagger} ) \right]
    & = \sum_{r,s} f_{rs} [ \Psiop, (\Psiop^{\prime\dagger})^r (\Psiop^\prime)^s ] \\
    & = \sum_{r,s} f_{rs} [ \Psiop, (\Psiop^{\prime\dagger})^r ] (\Psiop^\prime)^s \\
    & = \sum_{r,s} f_{rs} r \delta_P(\xvec^\prime, \xvec)
        (\Psiop^{\prime\dagger})^{r-1} (\Psiop^\prime)^s \\
    & = \delta_P (\xvec^\prime, \xvec) \frac{\partial f}{\partial \Psiop^{\prime\dagger}}.
    \qedhere
\end{eqn}
\end{proof}
