% =============================================================================
\chapter{Squeezing in BEC interferometry}
\label{cha:bec-squeezing}
% =============================================================================

As we have seen in the previous chapters, the truncated Wigner method provides significant corrections to such important experimental observables as visibility or phase noise.
While they do establish the theoretical limits for these quantities, real measurements are still mostly limited by technical noise, which do not require quasiprobability methods to account for.

In this chapter we will consider a property of \abbrev{bec} interferometry experiments which cannot be obtained from the mean-field approximation.
This property is the degree of spin squeezing, which allows to reduce the minimum uncertainty of the collective spin of a two-component \abbrev{bec} below the shot noise limit.
A state with a high degree of coherence between the two components, the coherent spin state~\cite{Kitagawa1993}, has equal uncertainty in any direction.
Further evolution and nonlinear interactions can redistribute quantum noise, making it possible to achieve an uncertainty lower than the shot noise along one direction (at the price of the increased uncertainty in other directions, to preserve the Heisenberg inequality).
Spin squeezing has been theoretically introduced by Kitagawa and Ueda~\cite{Kitagawa1993}, and by Wineland \textit{et~al}~\cite{Wineland1994}, and later observed experimentally~\cite{Hald1999,Kuzmich2000}.
It allows an outcome of interferometry experiments to be measured with uncertainty beyond the standard shot noise limit~\cite{Riedel2010,Gross2010} and can serve as a measure of entanglement in a two-component \abbrev{bec}~\cite{Sorensen2001,He2012-entanglement}.

In this chapter we will derive an expression that will allow us to obtain the degree of squeezing from the moments of wavefunctions in the truncated Wigner method, and apply it to a recent spin squeezing experiment.
We will also consider another possible experiment with a \abbrev{bec} near a Feshbach resonance where spin squeezing can be observed and discuss the complications caused by nonlinear losses.


\input{chapters/bec-squeezing/theory}
\input{chapters/bec-squeezing/separation}
\input{chapters/bec-squeezing/feshbach}
\input{chapters/bec-squeezing/conclusion}
