% =============================================================================
\chapter{Functional FPE to SDE correspondences}
\label{cha:appendix:fpe-sde}
% =============================================================================

Wigner transformation, to which the majority of this thesis is dedicated to, produces a \abbrev{fpe}, or its functional equivalent, from an initial master equation.
\abbrev{fpe} is an equation in partial derivatives, and, in general, is not easy to solve~--- even numerically.
The major part of the usefullness of the Wigner transformation paired with the Wigner truncation is that it produces \abbrev{fpe} in a special form, which can be further transformed to a set of \abbrev{sde}s, with the Wigner function playing the role of a probability distribution.
Algorithms of solving such equations numerically are much more straightforward.

The actual correspondence between \abbrev{fpe} and \abbrev{sde}s is formulated and proved for real-valued coefficients in literature~\cite{Risken1996}.
In this thesis we will need to transform \abbrev{fpe}s with complex coefficients, or even functional operator ones.
While it is always possible to express them in real-valued form, it is much more convenient to derive correspondence theorems that work directly on such \abbrev{fpe}s.
In this Appendix we will do that by proceeding successively from the initial real-valued theorem to complex-valued and functional correspondences.

In addition, we will do the same for the It\^o formula, which provides the expression for the time derivative of any function of transverse variables.
This formula is useful, among other cases, if one wants to derive the time dependence of some integral observable (for instance, population), without solving \abbrev{sde}s themselves.
Alternatively, it can serve as an additional test of a numerical algorithm used to propagate \abbrev{sde}s in time.

\input{chapters/fpe-sde/corr}
\input{chapters/fpe-sde/ito-formula}
